On exponential sums arising from the classical doubling method

### 2025-03-24 reorganization.tex
Calvin update

\subsubsection{Rational Tori}
A \emph{rational torus} $\algebraicGroup{T} \subset \algebraicGroup{G}$ over $\finiteField$ is a torus defined over $\finiteField$. 
Over a finite field, $\algebraicGroup{G}(\finiteField)$ conjugacy classes of maximal rank rational tori \calvin{check if we need connected or something} are parameterized by  Frobenius order twisted conjugacy classes in the Weyl group. In particular if you fix a rational torus $\algebraicGroup{T}_0$ with rank $=\mathrm{rk} \algebraicGroup{G}$ and an element $g \in \algebraicGroup{G}(\algebraicClosure{\finiteField})$ such that $g^{-1}\Frobenius(g) \in N(\algebraicGroup{T}_0)$ then the torus $g\algebraicGroup{T}_0g^{-1}$ is rational, and it's $\algebraicGroup{G}(\finiteField)$-conjugacy class is determined  by the Frobeinus-twisted conjugacy class of the image of $g^{-1}\Frobenius(g)$ in $W(\algebraicGroup{T}_0)(\algebraicClosure{\finiteField})$ \cite[Section 3.3]{Carter1985}.

@@

Let $\algebraicGroup{A}_n \subset \algebraicGroup{\GL}_n$ be the diagonal torus. We denote for an element $g \in \algebraicGroup{\GL}_n$, $g^{\ast} = w_n \transpose{g}^{-1} w_n$, where $w_n$ is the long Weyl element (see \Cref{subsec:isometry-groups}). Using $\algebraicGroup{A}_n$ we define standard (mostly split) tori in our groups as follows.
\begin{center}
\begin{tabular}{|c|c|}
	\hline 
		$\algebraicGroup{G}$ & $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$ \tabularnewline
		\hline 
		\hline 
		$\algebraicGroup{\UnitaryGroup}_{2n}$ & $\left\{ \begin{pmatrix}
			a\\
			& \involution{\left(a^{\ast}\right)}
		\end{pmatrix}  \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
		\hline 
		$\algebraicGroup{\UnitaryGroup}_{2n+1}$ & $\left\{ \begin{pmatrix}
	a \\
	& t\\
	& & \involution{\left(a^{\ast}\right)}
\end{pmatrix}  \mid a \in \algebraicGroup{A}_n,\,\, t \in \algebraicGroup{U}_1 \right\}$ \tabularnewline
\hline 
	$\algebraicGroup{\SO}_{2n+1}$ & $\left\{ \begin{pmatrix}
		a\\
		& 1\\
		& & a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
	\hline 
	$\algebraicGroup{\Sp}_{2n}$ or $\algebraicGroup{\SO}^{+}_{2n}$ & $\left\{ \begin{pmatrix}
		a\\
		& a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
	\hline 
	$\algebraicGroup{\SO}_{2n}^{-}$ & $\left\{ \begin{pmatrix}
		a\\
		& h\\
		& & a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_{n-1}\,\, h \in \algebraicGroup{\SO}_2^{-} \right\}$ \tabularnewline
	\hline		
	\end{tabular}
\end{center}
Here $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$, is split for all rows except for the the last row. We also write down the standard tori for the relevant corresponding similitude groups.
\begin{center}
	\begin{tabular}{|c|c|}
		\hline 
		$\algebraicGroup{G}$ & $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$ \tabularnewline
		\hline 
		\hline 
		$\algebraicGroup{\GSp}_{2n}$ or $\algebraicGroup{\GSO}^{+}_{2n}$ & $\left\{\left(\lambda, \begin{pmatrix}
			\lambda a\\
			& a^{\ast}
		\end{pmatrix}\right) \mid a \in \algebraicGroup{A}_n,\,\, \lambda \in \multiplcativeScheme \right\}$ \tabularnewline
		\hline 
		$\algebraicGroup{\GSO}_{2n}^{-}$ & $\left\{ \left(\algebraicGroup{\aFieldNorm}_{\finiteFieldExtension{2} \slash \finiteField}\left(h\right), \begin{pmatrix}
			\algebraicGroup{\aFieldNorm}_{\finiteFieldExtension{2} \slash \finiteField}\left(h\right) a\\
			& h\\
			& & a^{\ast}
		\end{pmatrix}\right) \mid a \in \algebraicGroup{A}_{n-1}\,\, h \in \restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme} \right\}$ \tabularnewline
		\hline		
	\end{tabular}
\end{center}
Here we realize $\restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme}$ as $$\restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme} = \left\{ \left(\lambda, \begin{pmatrix}
	a & b\\
	db & a
\end{pmatrix}\right) \mid a^2 - db^2 = \lambda \in \multiplcativeScheme \right\},$$
where $d$ is as in \Cref{subsec:isometry-groups}.

Applying this theory to our groups $\algebraicGroup{G}$
\begin{enumerate}
	\item (Special orthogonal groups): $\quadraticExtension = \finiteField$ and $\epsilon_{\hermitianSpace} = 1$. It follows by definition that every element $g \in \IsometryGroup\left(\hermitianSpace\right)$ satisfies $\left(\det g\right)^2 = 1$.  In this case, we obtain the connected components of the identity by adding a restriction on the determinant: We have that $G$ consists of all elements $g \in \IsometryGroup\left(\hermitianSpace\right)$ with $\det g = 1$. There are two cases here.
	\begin{enumerate}
		\item Split case: in this case, we can write $$\hermitianSpace = \begin{dcases}
		\xIsotropic \oplus \yIsotropic & \dim_{\finiteField} \hermitianSpace = 2n\\
		\xIsotropic \oplus \finiteField v \oplus \yIsotropic & \dim_{\finiteField}\hermitianSpace = 2n + 1
		\end{dcases}$$ where $\xIsotropic$ and $\yIsotropic$ are totally isotropic (that is, for every $x \in \xIsotropic$, $\innerproduct{x}{x} = 0$ and similarly for $\yIsotropic$) and these spaces are in duality with respect to $\innerproduct{\cdot}{\cdot}$, i.e., the map $\yIsotropic \to \Hom_{\finiteField}\left(\xIsotropic, \finiteField\right)$ given by $y \mapsto \left(x \mapsto \innerproduct{x}{y}\right)$ is an isomorphism. In the odd dimensional case, $v \in \hermitianSpace$ is an element orthogonal to $\xIsotropic$ and $\yIsotropic$ that satisfies $\innerproduct{v}{v} \ne 0$.

		In either case, denote $d = \dim_{\finiteField} \hermitianSpace$. We have that $G$ is isomorphic to $\SO^{+}_{d}\left(\finiteField\right) \subset  \GL_{d}\left(\finiteField\right)$, defined as the $\finiteField$-points of $$\algebraicGroup{\SO}^{+}_{d} = \left\{g \in \algebraicGroup{\GL}_{d} \mid \transpose{g} w_{d} g  = w_{d},\,\, \det g = 1\right\}.$$

		In the odd-dimensional case $d = 2n+1$, we are always in this situation, and we simply denote $\SO_{2n+1}\left(\finiteField\right) = \SO_{2n+1}^{+}\left(\finiteField\right)$ and $\algebraicGroup{\SO}_{2n+1} = \algebraicGroup{\SO}_{2n+1}^{+}$.
		\item Non-split case: In this case, $\dim_{\finiteField} \hermitianSpace = 2n$ and there exists a decomposition $$\hermitianSpace = \xIsotropic \oplus \hermitianSpace' \oplus \yIsotropic,$$
		where $\xIsotropic$ and $\yIsotropic$ are totally isotropic spaces in duality and $\hermitianSpace'$ is an antisotropic space (that is, for any $0 \ne v' \in \hermitianSpace'$, we have $\innerproduct{v'}{v'} \ne 0$) of dimension 2. Let $d$ be an element of $\multiplicativegroup{\finiteField}$ that is not a square. Denote $$B_{n,d} = \begin{pmatrix}
			\IdentityMatrix{n-1}\\
			& & 1 &\\
			& -d & &\\
			& & & \IdentityMatrix{n-1}
		\end{pmatrix} \cdot w_{2n}.$$
		In this case, we have that $G$ is isomorphic to $\SO^{-}_{2n}\left(\finiteField\right) \subset  \GL_{2n}\left(\finiteField\right)$, defined as the $\finiteField$-points of $$\algebraicGroup{\SO}^{-}_{2n} = \left\{g \in \algebraicGroup{\GL}_{2n} \mid \transpose{g} B_{n,d} g = B_{n,d},\,\, \det g = 1\right\}.$$
	\end{enumerate}
	\item (Symplectic groups): $\quadraticExtension = \finiteField$ and $\epsilon_{\hermitianSpace} = -1$. In this case, $\dim_{\finiteField} \hermitianSpace = 2n$ for some positive integer $n$ and $G$ is isomorphic to the group $\Sp_{2n}\left(\finiteField\right) \subset  \GL_{2n}\left(\finiteField\right)$, defined as the $\finiteField$-points of $$\algebraicGroup{\Sp}_{2n} = \left\{g \in \algebraicGroup{\GL}_{2n} \mid \transpose{g} \begin{pmatrix}
		& w_n\\
		-w_n
	\end{pmatrix} g = \begin{pmatrix}
		& w_n\\
		-w_n
	\end{pmatrix}\right\}.$$
	\item (Unitary groups): $\quadraticExtension \ne \finiteField$. Denote $\dim_{\quadraticExtension} \hermitianSpace = n$. In this case, the group $G = \GroupExtension{G}$ is isomorphic to the group $\UnitaryGroup_n\left(\finiteField\right) \subset\GL_{n}\left(\quadraticExtension\right)$ defined as the $\finiteField$-points of $$\algebraicGroup{\UnitaryGroup}_n = \left\{ g \in \restrictionOfScalars{\quadraticExtension}{\finiteField}{\algebraicGroup{\GL}_n} \mid \involution{\left(\transpose{g}\right)} w_n g = w_n \right\}.$$
\end{enumerate}

>>>>

\subsubsection{Rational Tori}
A \emph{rational torus} $\algebraicGroup{T} \subseteq \algebraicGroup{G}$ over $\mathbb{F}$ is a torus defined over $\mathbb{F}$. 
Over a finite field, $\algebraicGroup{G}(\mathbb{F})$ conjugacy classes of maximal rank rational tori \calvin{check if we need connected or something} are parameterized by  Frobenius order twisted conjugacy classes in the Weyl group. In particular if you fix a rational torus $\algebraicGroup{T}_0$ with rank $=\mathrm{rk}(\algebraicGroup{G}$ and an element $g \in \algebraicGroup{G}(\algebraicClosure{\mathbb{F}})$ such that $g^{-1}\Frobenius(g) \in N(\algebraicGroup{T}_0)$ then the torus $g\algebraicGroup{T}_0g^{-1}$ is rational, and it's $\algebraicGroup{G}(\mathbb{F})$-conjugacy class is determined  by the Frobeinus-twisted conjugacy class of the image of $g^{-1}\Frobenius(g)$ in $W(\algebraicGroup{T}_0)(\algebraicClosure{\mathbb{F}})$ [cite Roger Carter Finite Groups of Lie Type section 3.3].%since $\Frobenius(g\algebraicGroup{T}_0g^{-1}) = g\algebraicGroup{T}_0g^{-1}$
%could really demonstrate this by stating Land map is surjective + that all maximal tori are conjugate over an algebraiclly closed field
Applying this theory to our groups $\algebraicGroup{G}$, for each group we find a nice rational torus $\algebraicGroup{T}$ and write out the Frobenius order action on the Weyl group and the

\begin{enumerate}
	\item(Special orthogonal groups) 
	In each case let $\algebraicGroup{T}$ be the diagonal torus. Then $W(\algebraicGroup{T} )$ has representatives the type B or D permutation matrices which are fixed by the Frobenius order.
	\begin{enumerate}[(a)]
		\item (Split odd case $\algebraicGroup{SO}_{2n+1}$) $\algebraicGroup{T} = \algebraicGroup{G}_m^{n}$. In type B the Weyl group conjugacy classes are given by $(\lambda^+,\lambda^-)$ where $\lambda^+$ is the partition formed from and $\lambda^-$ is the partition formed from. Given a representation $w$ of the conjugacy class we get:
		$$^w\algebraicGroup{T} = \Pi_{\lambda^+_i} \mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}}\algebraicGroup{G}_m \times \Pi_{\lambda^-_i} \mathrm{Res}_{\mathbb{F}_{\lambda^-_i}:\mathbb{F}}\algebraicGroup{U}(1)$$
		\item (Split even case $\algebraicGroup{SO}_{2n}^+$) $\algebraicGroup{T} = \algebraicGroup{G}_m^{n}$.  Given $w \in W(\algebraicGroup{T})$ the rational structure of $^w\algebraicGroup{T}$ is determined by the conjugacy class of $w$ in the Type B Weyl group (a $\mathbb{Z}/2\mathbb{Z}$ extension of $W(\algebraicGroup{T})$. If $w$ is conjugate in the extension to $(\lambda^+,\lambda^-)$ where $\lambda^+$ is the partition formed from and $\lambda^-$ is the partition formed from:
		$$^w\algebraicGroup{T} = \Pi_{\lambda^+_i} \mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}}\algebraicGroup{G}_m \times \Pi_{\lambda^-_i} \mathrm{Res}_{\mathbb{F}_{\lambda^-_i}:\mathbb{F}}\algebraicGroup{U}(1)$$
			\item (Non-split even case $\algebraicGroup{SO}_{2n}^-$) Set $\algebraicGroup{T} = \algebraicGroup{G}_m^{n-1} \times \algebraicGroup{U}(1)$ the diagonal torus. Given $w \in W(\algebraicGroup{T})$ the rational structure of $^w\algebraicGroup{T}$ is determined by the conjugacy class of $w$ in the Type B Weyl group (a $\mathbb{Z}/2\mathbb{Z}$ extension of $W(\algebraicGroup{T})$. If $w \circ (-1,1)$ is conjugate in the extension to $(\lambda^+,\lambda^-)$ where $\lambda^+$ is the partition formed from and $\lambda^-$ is the partition formed from:
		$$^w\algebraicGroup{T} = \Pi_{\lambda^+_i} \mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}}\algebraicGroup{G}_m \times \Pi_{\lambda^-_i} \mathrm{Res}_{\mathbb{F}_{\lambda^-_i}:\mathbb{F}}\algebraicGroup{U}(1)$$
	\end{enumerate}
	\item (Symplectic groups) In type B the Weyl group conjugacy classes are given by $(\lambda^+,\lambda^-)$ where $\lambda^+$ is the partition formed from and $\lambda^-$ is the partition formed from. Given a representation $w$ of the conjugacy class we get:
		$$^w\algebraicGroup{T} = \Pi_{\lambda^+_i} \mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}}\algebraicGroup{G}_m \times \Pi_{\lambda^-_i} \mathrm{Res}_{\mathbb{F}_{\lambda^-_i}:\mathbb{F}}\algebraicGroup{U}(1)$$
	\item (Unitary groups) We let our torus be the diagonal torus for the unitary group with hermitian form $I_n$. Then $\algebraicGroup{T}=U(1)^n$ and the Frobenius action on the Weyl group is trivial. Note since all hermitian forms are equivalent over a finite field, $\algebraicGroup{T}$ is a torus in the group $\algebraicGroup{U}_n$ from the previous section.  In type A the Weyl group conjugacy classes are given by $(\lambda)$, and
	$$^w\algebraicGroup{T} = \Pi_{\lambda_i} \begin{cases}
	\mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}} \algebraicGroup{G}_m& \lambda_i \textnormal{ is even} \\
	\mathrm{Res}_{\mathbb{F}_{\lambda^+_i}:\mathbb{F}}\algebraicGroup{U}(1) & \lambda_i \textnormal{ is odd}
	\end{cases}$$
	\item (Even special orthogonal similitude groups)
	\item (Symplectic similitude groups)
\end{enumerate}

@@

Let $\algebraicGroup{A}_n \subset \algebraicGroup{\GL}_n$ be the diagonal torus. We denote for an element $g \in \algebraicGroup{\GL}_n$, $g^{\ast} = w_n \transpose{g}^{-1} w_n$, where $w_n$ is the long Weyl element (see \Cref{subsec:isometry-groups}). Using $\algebraicGroup{A}_n$ we define standard (mostly split) tori in our groups as follows.
\begin{center}
\begin{tabular}{|c|c|}
	\hline 
		$\algebraicGroup{G}$ & $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$ \tabularnewline
		\hline 
		\hline 
		$\algebraicGroup{\UnitaryGroup}_{2n}$ & $\left\{ \begin{pmatrix}
			a\\
			& \involution{\left(a^{\ast}\right)}
		\end{pmatrix}  \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
		\hline 
		$\algebraicGroup{\UnitaryGroup}_{2n+1}$ & $\left\{ \begin{pmatrix}
	a \\
	& t\\
	& & \involution{\left(a^{\ast}\right)}
\end{pmatrix}  \mid a \in \algebraicGroup{A}_n,\,\, t \in \algebraicGroup{U}_1 \right\}$ \tabularnewline
\hline 
	$\algebraicGroup{\SO}_{2n+1}$ & $\left\{ \begin{pmatrix}
		a\\
		& 1\\
		& & a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
	\hline 
	$\algebraicGroup{\Sp}_{2n}$ or $\algebraicGroup{\SO}^{+}_{2n}$ & $\left\{ \begin{pmatrix}
		a\\
		& a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_n \right\}$ \tabularnewline
	\hline 
	$\algebraicGroup{\SO}_{2n}^{-}$ & $\left\{ \begin{pmatrix}
		a\\
		& h\\
		& & a^{\ast}
	\end{pmatrix} \mid a \in \algebraicGroup{A}_{n-1}\,\, h \in \algebraicGroup{\SO}_2^{-} \right\}$ \tabularnewline
	\hline		
	\end{tabular}
\end{center}
Here $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$ is split for all rows except for the the last row.
Here $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$, is split for all rows except for the the last row. We also write down the standard tori for the relevant corresponding similitude groups.
\begin{center}
	\begin{tabular}{|c|c|}
		\hline 
		$\algebraicGroup{G}$ & $\algebraicGroup{T}^{\mathrm{std}}_{\algebraicGroup{G}}$ \tabularnewline
		\hline 
		\hline 
		$\algebraicGroup{\GSp}_{2n}$ or $\algebraicGroup{\GSO}^{+}_{2n}$ & $\left\{\left(\lambda, \begin{pmatrix}
			\lambda a\\
			& a^{\ast}
		\end{pmatrix}\right) \mid a \in \algebraicGroup{A}_n,\,\, \lambda \in \multiplcativeScheme \right\}$ \tabularnewline
		\hline 
		$\algebraicGroup{\GSO}_{2n}^{-}$ & $\left\{ \left(\algebraicGroup{\aFieldNorm}_{\finiteFieldExtension{2} \slash \finiteField}\left(h\right), \begin{pmatrix}
			\algebraicGroup{\aFieldNorm}_{\finiteFieldExtension{2} \slash \finiteField}\left(h\right) a\\
			& h\\
			& & a^{\ast}
		\end{pmatrix}\right) \mid a \in \algebraicGroup{A}_{n-1}\,\, h \in \restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme} \right\}$ \tabularnewline
		\hline		
	\end{tabular}
\end{center}
Here we realize $\restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme}$ as $$\restrictionOfScalars{\finiteFieldExtension{2}}{\finiteField}{\multiplcativeScheme} = \left\{ \left(\lambda, \begin{pmatrix}
	a & b\\
	db & a
\end{pmatrix}\right) \mid a^2 - db^2 = \lambda \in \multiplcativeScheme \right\},$$
where $d$ is as in \Cref{subsec:isometry-groups}.
